---
title: "STA6990: Week 2"
author: "Chantal Ojurongbe"
execute:
  echo: true
  warning: false
  message: false
format: 
  html:
    embed-resources: true
editor: source
---

**1. You drive to work 20 days, keeping track of the commuting time (in minutes) for each trip. Here are the twenty measurements.**

<center>
25.4, 27.8, 26.8, 24.1, 24.5, <br>
23.0, 27.5, 24.3, 28.4, 29.0, <br>
29.4, 24.9, 26.3, 23.5, 28.3, <br>
27.8, 29.4, 25.7, 24.3, 24.2 
</center>
<br>
**a. Approximate the probability that it takes you under 25 minutes to drive to work.**
```{r}
commute_times <- c(25.4, 27.8, 26.8, 24.1, 24.5, 23.0, 27.5, 24.3, 28.4, 29.0, 
                   29.4, 24.9, 26.3, 23.5, 28.3, 27.8, 29.4, 25.7, 24.3, 24.2)

days_under_25 <- sum(commute_times < 25)
total_days <- length(commute_times)
probability_under_25 <- days_under_25 / total_days
probability_under_25
```
The probability that it takes you under 25 minutes to drive to work is 40%.

**b. Approximate the probability it takes between 25 and 28 minutes to drive to work.**
```{r}
commute_times <- c(25.4, 27.8, 26.8, 24.1, 24.5, 23.0, 27.5, 24.3, 28.4, 29.0, 
                   29.4, 24.9, 26.3, 23.5, 28.3, 27.8, 29.4, 25.7, 24.3, 24.2)

days_between_25_and_28 <- sum(commute_times >= 25 & commute_times <= 28)
total_days <- length(commute_times)
probability_between_25_and_28 <- days_between_25_and_28 / total_days
probability_between_25_and_28
```
The probability it takes between 25 and 28 minutes to drive to work is 35%

**c. Suppose one day it takes you 23 minutes to get to work. Would you consider this unusual? Why?**
```{r}
commute_times <- c(25.4, 27.8, 26.8, 24.1, 24.5, 23.0, 27.5, 24.3, 28.4, 29.0, 
                   29.4, 24.9, 26.3, 23.5, 28.3, 27.8, 29.4, 25.7, 24.3, 24.2)

mean_time <- mean(commute_times)
std_deviation <- sd(commute_times)

mean_time
std_deviation
```
We can tell whether a 23-minute commute is within two standard deviations of the mean. If it is, it's not unusual; if it's not, then it might be considered unusual. We look at the range defined by the mean plus or minus two standard deviations. This range covers approximately 95% of the data in a normal distribution. Range:

Lower bound = Mean - 2 * Standard Deviation
Upper bound = Mean + 2 * Standard Deviation
Lower bound = 26.23 - 2 * 2.09463
Upper bound = 26.23 + 2 * 2.09463

The range, based on two standard deviations from the mean, is approximately 22.04 to 30.42 minutes. Since a 23-minute commute falls within this range, it would not be considered unusual based on your commuting data.

**2. Suppose you roll three fair dice.**

**a. How many possible outcomes are there?**

The total number of possible outcomes when rolling three fair dice is 216.

**b. Find the probability you roll three fives.**
```{r}
favorable_outcomes <- 1
total_outcomes <- 6^3
probability_rolling_fives <- favorable_outcomes / total_outcomes
probability_rolling_fives
```
The probability you roll three fives is 0.46%.

**c. Find the probability that all three dice show the same number.**
```{r}
favorable_outcomes <- 6  
total_outcomes <- 6^3
probability_same_number <- favorable_outcomes / total_outcomes
probability_same_number
```
The probability that all three dice show the same number is 2.78%

**d. Find the probability that the sum of the dice is equal to 10.**
```{r}
favorable_outcomes <- 0

for(die1 in 1:6) {
  for(die2 in 1:6) {
    for(die3 in 1:6) {
      if(die1 + die2 + die3 == 10) {
        favorable_outcomes <- favorable_outcomes + 1}}}}

total_outcomes <- 6^3
probability_sum_10 <- favorable_outcomes / total_outcomes
probability_sum_10
```
The probability that the sum of the dice is equal to 10 is 12.5%.

**3. CHALLENGE. The built-in vector, `letters` contains the 26 lower-case letters of the alphabet.**

**a. Using the `sample()` function, take a sample of 10 letters without replacement from `letters`.**
```{r}
sampled_letters <- sample(letters, 10, replace = FALSE)
sampled_letters
```

**b. Using the `sample()` function, take a sample of 10 letters with replacement from `letters`.**
```{r}
sampled_letters_with_replacement <- sample(letters, 10, replace = TRUE)
sampled_letters_with_replacement
```

