---
title: "STA6990: Week 1"
author: "Chantal Ojurongbe"
execute:
  echo: true
  warning: false
  message: false
format: 
  html:
    embed-resources: true
editor: source
---

<!-- This is a HTML comment. It will not appear in the rendered file. -->

```{r example}
# this is a R comment. it will show in the rendered file when the chunk is set to display
# for this file, we have the default chunk option set to display (controlled by the `echo` option in the yaml)

# this differently colored section is called a code chunk
# we can add code chunks using the green button with a C on it in the toolbar
# we can also add code chunks using a keyboard shortcut.

# at the beginning of the chunk, we told the file that it needs to use R to analyze what is here
# we also gave the chunk a name; you can see the chunks and their names at the bottom of this panel
```

```{r no_show}
#| echo: false # this is a chunk option
# by setting echo to false, we are telling Quarto to suppress this code chunk in the output
```

```{r setup}
library(tidyverse)
# if the above line throws an error, remove the # from the next line and run it
# install.packages("tidyverse")
# once library(tidyverse) works, delete the commented section here
# leaving certain commands in a .qmd file will keep it from rendering properly
# view(dataframe) is another one that will prevent rendering
```

**1. What is your name? Change the font color to be your favorite color. Note: this is asking for you to write HTML code. It is *not* R code.**

<font color="#9900FF">
Chantal Ojurongbe
</font>

**2. Is there a nickname you go by that you would like to go by in the classroom? Change the font to be Italicized.**

<em>Chantal</em>

**3. What are your career goals after you are finished with the program? Change the font to be underlined.**

<u>I plan to take the actuary first test and see how I fair. Depending on how that goes, I will either continue to study for the other test or get the data science certificate and progress my career in project management. At some point I will continue on to get my PHd, after about a semester of break.</u>

**4. What questions do you have about this course? Write them using an enumerated list.**

<ol type="1">
  <li>I have never heard of Bayesian anything. How is this used in a practical sense?</li>
  <li>I am glad we are learning how to code in html, but why?</li>
  <li>Will this be helpful in running analysis for PHD work. Will I be able to run my own analysis?</li>
</ol>

**5. What is the keyboard shortcut to render this file?**

The keyboard shortcut Ctrl + Shift + K

**6. Render this file using the keyboard shortcut and tell me where the resulting .html file saved to.**

~/STA6990

**7. What is the keyboard shortcut to add a code chunk?**

The keyboard shortcut Ctrl + Alt + I

```{r code chunk}
# Code chunk added.
```


**8. What is the keyboard shortcut to run a single line within a code chunk?**

Select the line and then keyboard shortcut Ctrl + Enter

**9. Import the dataset located on Canvas.**

```{r sas dataset}
library(haven)

sas_data <- read_sas("analysis1 (2).sas7bdat")
head(sas_data)
```

**10. Find an interesting (to you!) dataset from [Tidy Tuesday](https://github.com/rfordatascience/tidytuesday) and import it here.**

```{r csv dataset}
data1csv <- read_csv("recent-grads.csv", show_col_types = FALSE)
print(colnames(data1csv))
head(data1csv)
```

**11. Use functions within `tidyverse` to find an appropriate summary statistic for one variable in the dataset. Hint: Dr. Google is your friend! Try searching "summary statistic tidyverse".**

```{r summary stats}
library(tidyverse)

summary_stats <- data1csv %>%
  summarise(
    Mean_Median = mean(Median, na.rm = TRUE),
    Median_Median = median(Median, na.rm = TRUE),
    Sum_Median = sum(Median, na.rm = TRUE),
    Min_Median = min(Median, na.rm = TRUE),
    Max_Median = max(Median, na.rm = TRUE)
  )

print(summary_stats)
```

**12. Bonus! Use `ggplot()` to construct a visualization relevant to the dataset.**

```{r graph}
ggplot(data1csv, aes(x = Median)) +
  geom_histogram(binwidth = 3000, fill = "blue", color = "green") +
  theme_minimal() +
  labs(title = "Distribution of Median Salaries",
       x = "Median Salary",
       y = "Frequency")
   
```

