---
title: "STA6990: Week 3"
author: "Chantal Ojurongbe"
execute:
  echo: true
  warning: false
  message: false
format: 
  html:
    embed-resources: true
editor: source
---
```{r}
library(glmnet)
library(nnet)
library(boot)
library(fastDummies)
library(haven)
library(lindia)
library(tidyverse)
library(gsheet)
library(DescTools)
library(FSA)
library(dplyr)
library(ggplot2)
library(car)
library(agricolae)
library(gmodels)
library(dplyr)
library(car)
library(readr)
library(broom)
library(modelr)
library(performance)
library(GGally)
library(emmeans)
library(visreg)
library(see)
library(patchwork)
library(caret)
```

**1. Suppose we roll two fair dice and you keep track of the larger of the two rolls.**

**a. What is the sample space?**

The sample space for the larger of the two dice rolls is {1, 2, 3, 4, 5, 6}
    
**b. Let $X$ = the larger of the two numbers. Derive the probability distribution.**

X = 1: Since each die has 6 faces, and there are 2 dice, there are 6×6=36 6×6=36 possible outcomes).
X = 2: (1,2), (2,1), and (2,2). That's 3 out of 36.
X = 3: (1,3), (2,3), (3,1), (3,2), and (3,3). That's 5 out of 36.
X = 4: (1,4), (2,4), (3,4), (4,1), (4,2), (4,3), and (4,4). That's 7 out of 36.
X = 5: (1,5), (2,5), (3,5), (4,5), (5,1), (5,2), (5,3), and (5,4), and (5,5). That's 9 out of 36.
X = 6: (1,6), (2,6), (3,6), (4,6), (5,6), (6,1), (6,2), (6,3), (6,4), (6,5), and (6,6). That's 11 out of 36.

P(X=1)= 1/36  ≈0.028
P(X=2)= 3/36  ≈0.083
P(X=3)= 5/36  ≈0.139
P(X=4)= 7/36  ≈0.194
P(X=5)= 9/36  =0.25
P(X=6)= 11/36 ≈0.306
This probability distribution shows that it's more likely for the larger number to be higher, with the probability increasing as the value of X increases from 1 to 6.

**c. Verify that your answer in (b) is a valid probability distribution.**

To verify that the derived probability distribution is valid, it must satisfy two key conditions:
Non-Negativity: Each probability value P(X=x) must be non-negative (i.e., P(X=x)≥0 for all x).
Total Probability Equals 1: The sum of all the probability values must equal 1 (i.e., ∑P(X=x)=1).
P(X=1)= 1/36
P(X=2)= 3/36
P(X=3)= 5/36
P(X=4)= 7/36
P(X=5)= 9/36
P(X=6)= 11/36
All these probabilities are non-negative. The sum of all the probability values in the distribution is 1.0. Since each probability value is non-negative and their sum equals 1, the probability distribution is valid.

**2. A man with 5 keys wants to open the door and tries keys at random. Exactly one key will open the door. Assume that unsuccessful keys are eliminated from future attempts.**

**a. What is the sample space?**

The sample space is: { {1}, {2}, {3}, {4}, {5} }, where each number is the attempts where the correct key is used. This shows all the possible outcomes for the number of attempts needed to open the door.

**b. Let $X$ = the key's number that opens the door. Derive the probability distribution.**

The probability distribution for X is:
P(X=1)= 1/5 =0.2
P(X=2)= 1/5 =0.2 
P(X=3)= 1/5 =0.2
P(X=4)= 1/5 =0.2
This distribution is valid as each probability is non-negative and the sum of all probabilities equals.

**c. Verify that your answer in (b) is a valid probability distribution.**

To verify that the derived probability distribution is valid, it must satisfy two key conditions:
Non-Negativity: Each probability value P(X=x) must be non-negative (i.e., P(X=x)≥0 for all x).
Total Probability Equals 1: The sum of all the probability values must equal 1 (i.e., ∑P(X=x)=1).
P(X=1) =0.2
P(X=2) =0.2 
P(X=3) =0.2
P(X=4) =0.2
All these probabilities are non-negative. The sum of all the probability values in the distribution is 1.0. Since each probability value is non-negative and their sum equals 1, the probability distribution is valid.

**d. Take the code in lecture used to simulate values and edit to simulate 1000 trials. Graph the resulting distribution.**

```{r}
one_trial <- function() {
  keys <- 1:5
  sample(keys, size = 1, replace = TRUE)
}

set.seed(8571) 
sim_1000 <- tibble(Key = replicate(1000, one_trial()))

freq_distribution <- sim_1000 %>% count(Key)

freq_distribution %>% 
  ggplot(aes(x = Key, y = n)) +
  geom_col() + 
  labs(x = "Key Number", y = "Frequency", title = "Frequency Distribution of the Correct Key in 1000 Trials") + 
  scale_x_continuous(breaks = 1:5) + 
  theme_minimal()
```

**3. A standard drug is known to be effective in 80% of the cases in which it is used. A new drug is tested on 100 patients and found to be effective in 85 cases. Is the new drug superior? (Hint: Use R to find the probability of observing 85 or more successes assuming that the new and old drugs are equally effective.)**

```{r}
n <- 100         
p <- 0.8         
successes <- 85  

prob <- 1 - pbinom(successes - 1, n, p)
prob
```
The probability of observing 85 or more successes assuming that the new and old drugs are equally effective is 12.9%.

**4. CHALLENGE! Let the number of chocolate chips in a certain type of cookie have a Poisson distribution. Suppose that, on average, there are 7 chocolate chips per cookie.**

**a. Use R to find the probability of two chocolate chips in a cookie.**

```{r}
lambda <- 7  
k <- 2       

prob <- dpois(k, lambda)
prob
```
The probability of two chocolate chips in a cookie is 2.23%.

**b. Use R to find the probability of at least two chocolate chips in a cookie.**

```{r}
lambda <- 7  

prob_0_or_1 <- ppois(1, lambda)  

prob_at_least_2 <- 1 - prob_0_or_1

prob_at_least_2
```
The probability of at least two chocolate chips in a cookie is 99.27%.

**c. Simulate the results from 10000 cookies and graph the resulting distribution.**

```{r}

num_cookies <- 10000  
lambda <- 7           

choc_chips <- rpois(num_cookies, lambda)

cookies_df <- data.frame(ChocChips = choc_chips)

ggplot(cookies_df, aes(x = ChocChips)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  labs(title = "Distribution of Chocolate Chips in 10000 Cookies",
       x = "Number of Chocolate Chips",
       y = "Frequency") +
  theme_minimal()
```

